---
title: "Auditing a body of literature"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Auditing a body of literature}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r setup}
library(instarreport)
```

The introductory vignette covers working with a single study. This one covers many. You assemble a set of studies, then ask what a
field reports, what it does not, and how that changes.

## Use cases

If you are writing a review, a meta-analysis, or
a methods paper, you probably want to say something about how the
animals in your corpus were handled, and at the moment that means
reading every methods section yourself and keeping notes in a
spreadsheet you invented. A shared item set means those notes are
comparable with everyone else's, and the arithmetic is done for you.

Beyond that, a good deal opens up, and most of it is not about reporting
at all. A completed sheet records *what was done*. The actual
temperatures, diets, acclimation periods, anaesthetics, killing methods.
A set of them is a structured record of husbandry and methods practice,
not just a tally of what got mentioned.

**What the field actually does.** Nobody has a good description of how
invertebrates are typically kept and handled in research. What is the
modal housing temperature for *Drosophila* behavioural work? How often
is anything used for analgesia in a procedure that would warrant it in a
vertebrate? What proportion of field-collected animals are released, and
on what reasoning? These are answerable questions that currently are not
answered, because the information is scattered through prose.

**Methods as moderators.** Housing density, photoperiod, acclimation and
handling regime all shape outcomes. Extracting them across a corpus
gives a meta-analyst covariates rather than unexplained heterogeneity,
and may explain why two labs asking the same question of the same
species disagree.

**Refinement, and whether it spreads.** A better practice appears from
time to time. A humane endpoint, an anaesthetic protocol, a less
damaging trap design. A structured record makes it possible to watch
whether it diffuses, how fast, and where it stalls.

**Change over time.** Both practice and reporting move. Scoring a
comparable sample at intervals, or either side of a policy change,
turns an impression into a measurement.

**Comparison across whatever unit you care about.** Taxon, subdiscipline,
lab versus field, question type, jurisdiction, funder, decade,
institution, journal. Norms differ along all of these, and the
differences are often the interesting result rather than a nuisance.

**Where the evidence is missing.** An item almost nobody reports is
either being overlooked or is something the field genuinely does not know
how to answer. Both are worth knowing. The first is an advocacy problem,
the second a research agenda.

**Evidence for policy.** Journals, funders and ethics bodies weighing a
requirement reasonably ask what it would cost and what it would change.
Coverage data answers that far better than a principle does.

**Teaching.** Handing students a folder of sheets and asking what is
missing, or what they would have done differently, is a fast way into
both research ethics and methods criticism.

None of this needs the whole literature, or a whole journal. A
well-chosen sample of one taxon, one technique or one question is enough
to say something.

## Compiling a corpus

Three realistic routes, in rough order of how common they will be.

### Manual scoring

Read the methods sections and record, for each paper and each item,
whether it was reported. This is the main path and will stay that way.
Most published work has no INSTAR sheet, and careful reading is still
the most reliable way to decide whether an item was genuinely addressed.

Record it as a wide table, one row per paper and one column per
`item_id`, then hand it to `audit_from_matrix()`. Section
[From a scoring table](#from-a-scoring-table) covers the mechanics.

Worth doing regardless of scale. Score a subset twice, or have a second
reader do it, and check agreement before trusting the totals. Reporting
is not always a clean yes or no, and the boundary cases are where two
readers diverge.

### Deposited sheets

Where authors have completed and deposited a sheet, the work is already
done and no reading is required. Point `read_instar()` at the folder.

This is the smallest number of studies today and will grow slowly. It is
also the reason the sheet is plain text rather than only a figure. A
deposited sheet can be read back, a deposited figure cannot.

### Assisted extraction

Large language models can be used to pull structured answers out of
methods text, and for a corpus of a few hundred papers that is a large
saving. It is also easy to do badly. If you go this way, then:

* **Validate against hand-scoring.** Score a random subset yourself and
  compare, item by item. Report the agreement. Without this you have
  numbers of unknown quality.
* **Give the model the item descriptions**, not just the item names.
  `instar_items$description` is written to be exactly this kind of
  prompt material.
* **Allow an "uncertain" response** as well as yes, no, and not
  applicable. Forcing a binary choice on ambiguous text manufactures
  confidence that is not there. Resolve the uncertain ones by hand.
* **Report what you did** in your methods. Which model and version, the
  prompt or rubric, how many papers were re-scored by hand, and the
  agreement you found. This is a methods choice like any other and
  should be reproducible.
* **Feed it the methods section**, not the whole paper. Precision
  improves and cost falls.

The output is the same wide table as manual scoring, so it reaches
`audit_from_matrix()` the same way. The survey reported in the INSTAR
paper was built this way, and its methods section documents the model,
the rubric, and the hand-validation.

Whichever route you take, the resulting object is the same and
everything downstream is identical.

## Reading deposited sheets

`read_instar()` takes a file, a directory, or a vector of either. Let's
build a small corpus so the examples run.

```{r}
corpus_dir <- file.path(tempdir(), "supplements")
dir.create(corpus_dir, showWarnings = FALSE)

make_sheet <- function(n_reported, title, journal, doi, housing) {
  # instar_set() takes item_id = value pairs. Building them programmatically
  # means assembling a named list and using do.call().
  ids <- instar_items$item_id[seq_len(n_reported)]
  filled <- stats::setNames(as.list(rep("Reported in the paper", length(ids))),
                            ids)
  filled$env_housing <- housing
  items <- do.call(instar_set, c(list(instar_template()), filled))

  rep <- instar_report(items, paper = list(
    title = title, authors = "Author et al. (2026)",
    journal = journal, doi = doi
  ))
  slug <- gsub("[^A-Za-z0-9]+", "-", doi)
  write_report(rep, file.path(corpus_dir, paste0(slug, ".csv")))
}

invisible(Map(make_sheet,
  n_reported = c(4, 9, 14, 6, 11),
  title      = paste("Study", 1:5),
  journal    = c("J Alpha", "J Alpha", "J Beta", "J Beta", "J Beta"),
  doi        = paste0("10.1234/study", 1:5),
  housing    = c("25 +/- 1 C, 12:12 L:D, 60% RH",
                 "23 C, 14:10 L:D, group housed at 20 per container",
                 "Ambient (18-26 C), natural photoperiod",
                 "25 C, 12:12 L:D, individually housed",
                 "27 +/- 0.5 C, 16:8 L:D, 70% RH")))

list.files(corpus_dir)
```

Now read them:

```{r}
corpus <- read_instar(corpus_dir)
corpus
```

A corpus is a named list of reports underneath, so `lapply()`,
`Filter()`, `[` and friends all work on it. `summary()` gives one row per
sheet:

```{r}
summary(corpus)[, c("study", "title", "journal", "reported", "percent_reported")]
```

### When a file will not read

Pointed at a real directory you will meet malformed files. A deleted
column, a semicolon-delimited export, a stray spreadsheet someone left
in the folder. `read_instar()` does not stop on those. It warns, skips
them, and records what went wrong.

```{r}
writeLines(c("this,is,not", "an,instar,sheet"),
           file.path(corpus_dir, "junk.csv"))

corpus <- suppressWarnings(read_instar(corpus_dir))
attr(corpus, "failed")
```

Getting the other five sheets plus a list of what to go and fix is more
useful than getting nothing.

### Framework versions

Every sheet records the version it was completed against. A corpus that
mixes versions is warned about, because coverage across them is not
comparable. An item missing from half the sheets because it did not
exist yet is not the same as an item those studies failed to report.

```{r}
table(summary(corpus)$version, useNA = "ifany")
```

If you do have a mix, split before auditing:

```{r, eval = FALSE}
s <- summary(corpus)
v1 <- corpus[s$version == "1.0"]
instar_audit(v1)
```

## From a scoring table {#from-a-scoring-table}

However you produced it, whether by reading papers yourself, a second
coder, or assisted extraction, the shape is one row per paper and one
column per `item_id`.

```{r}
scores <- data.frame(
  doi            = c("10.1/a", "10.1/b", "10.1/c", "10.1/d"),
  journal        = c("J Alpha", "J Alpha", "J Beta", "J Beta"),
  year           = c(2024, 2025, 2025, 2026),
  subjects_taxon = c("Y", "Y", "Y", "Y"),
  subjects_n     = c("Y", "N", "N", "Y"),
  env_housing    = c("Y", "Y", "NA", "N"),
  env_field      = c("NA", "NA", "Y", "-"),
  stringsAsFactors = FALSE
)

audit2 <- audit_from_matrix(scores, id = "doi")
summary(audit2)[, c("item", "reported", "applicable", "percent_reported")]
```

Three things there are worth spelling out.

**Cell values are read leniently**, because scoring sheets are made by
people. `Y`, `yes`, `TRUE`, `1` and `C` all mean reported. `N`, `no`,
`FALSE` and `0` mean not. `NA`, `N/A`, `-` and blanks mean not
applicable. Anything unrecognised warns rather than being silently
counted, so a stray `?` or `U` from an uncertain call will not slip
through as a decision you did not make.

**Non-item columns become study metadata.** `journal` and `year` are not
`item_id`s, so they ride along and are immediately available for
grouping:

```{r}
subset(summary(audit2, by = "year"), item_id == "subjects_n")
```

**Items absent from the table are left out**, not counted as unreported.
The message above says which. Recording an unscored item as a failure to
report would assert something the scoring never checked, which matters
if you scored a subset of items deliberately.

### Choosing the identifier

`id` should name a column that identifies every row. If it does not, you
get told rather than finding out later:

```{r}
incomplete <- data.frame(
  doi = c("10.1/a", "", ""),          # DOI extraction failed for two
  subjects_taxon = c("Y", "N", "Y"),
  stringsAsFactors = FALSE
)
audit3 <- suppressMessages(audit_from_matrix(incomplete, id = "doi"))
audit3$n
```

That warns about the blank ids and still counts three studies. It is a
real case. In the survey reported in the INSTAR paper, DOI extraction
failed for 25 of 150 PDFs, and keying on DOI would have collapsed them
into one repeated blank identifier. The fix was to key on a per-file
slug instead and keep DOI as metadata.

## Auditing

`instar_audit()` takes a corpus, a scoring table, or a path directly.

```{r}
audit <- instar_audit(corpus)
audit
```

`summary()` gives one row per framework item:

```{r}
head(summary(audit), 4)
```

Two columns are worth knowing. `applicable` is the denominator, which
excludes studies that marked the item not applicable, and
`percent_reported` is `reported / applicable`. An item that genuinely does not apply to a study
is never counted against it.

Which items does this literature handle worst?

```{r}
s <- summary(audit)
head(s[order(s$percent_reported), c("item", "reported", "applicable",
                                    "percent_reported")], 5)
```

### Grouping

Any study-level column can be a grouping variable. For a corpus read from
sheets that means anything the reserved rows carry, so `journal`, `doi`
and `version`, plus `folder`, the directory each sheet came from. For a
scoring table it means any column that was not an item.

```{r}
by_journal <- summary(audit, by = "journal")
subset(by_journal, item_id == "subjects_taxon")
```

`folder` is the useful one when a corpus is filed into directories that
mean something. Read recursively and the structure comes through:

```{r, eval = FALSE}
corpus <- read_instar("corpus/", recursive = TRUE)
summary(instar_audit(corpus), by = "folder")
```

### Per-study coverage

`audit$studies` is one row per study, which is what you want for
questions about papers rather than items:

```{r}
stats::aggregate(percent_reported ~ journal, data = audit$studies,
                 FUN = stats::median)
```

### Plotting

```{r, fig.width = 7, fig.height = 6, fig.alt = "Bar chart of reporting coverage per framework item"}
plot(audit)
```

`autoplot()` returns the ggplot object instead of drawing it, so you can
modify it before rendering. `plot(audit, by = "journal")` facets.

## Beyond coverage, to what the studies actually did

Coverage tells you whether an item was addressed. The sheet also carries
*what was said*, which for most of the questions in
[Use cases](#use-cases) is the part you want. Each report keeps the text
in its items table, so pulling one item across a corpus is a one-liner:

```{r}
d <- as.data.frame(corpus)

subset(d, item_id == "env_housing" & status == "reported",
       select = c(study, value))
```

From there it is ordinary text work. Parse temperatures out, tabulate
photoperiods, count how many housed animals individually. The framework
does not standardise the content, deliberately, since forcing husbandry
into fixed fields would lose more than it gained. It does guarantee that
the housing description sits in the housing row of every sheet, which is
most of the battle.

The same applies to any item. `fate_end` across a corpus tells you how
often animals are released, killed, or kept. `proc_anaesthesia` tells
you what is actually used, and how often nothing is. `ethics_review`
tells you how often review applied at all, and what reasoning was
offered when it did not.

`as.data.frame()` on a corpus gives one row per study per item, carrying
the paper's details alongside, so grouping and filtering are ordinary
data frame work.

```{r}
str(d, vec.len = 2)
```

```{r, include = FALSE}
unlink(corpus_dir, recursive = TRUE)
```
