---
title: "Auditing a body of literature"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Auditing a body of literature}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r setup}
library(instarreport)
```

The other vignette covers one study. This one covers many: reading a set
of completed sheets, or a set of scored papers, and asking what a field
reports and what it does not.

There are two ways in, depending on whether the studies have INSTAR
sheets of their own.

* **Prospective.** Studies deposit completed sheets, and a corpus is
  just a folder of them. Read them with `read_instar()`.
* **Retrospective.** The studies predate the framework, so someone reads
  the papers and scores them by hand into a table. Hand that to
  `audit_from_matrix()`.

Both produce the same `instar_audit` object, so everything downstream is
identical.

## Reading a folder of sheets

`read_instar()` takes a file, a directory, or a vector of either. Let's
build a small corpus to work with, so the examples run.

```{r}
corpus_dir <- file.path(tempdir(), "supplements")
dir.create(corpus_dir, showWarnings = FALSE)

make_sheet <- function(n_reported, title, journal, doi) {
  # instar_set() takes item_id = value pairs. Building them programmatically
  # means assembling a named list and using do.call().
  ids <- instar_items$item_id[seq_len(n_reported)]
  filled <- stats::setNames(as.list(rep("Reported in the paper", length(ids))),
                            ids)
  items <- do.call(instar_set, c(list(instar_template()), filled))

  rep <- instar_report(items, paper = list(
    title = title, authors = "Author et al. (2026)",
    journal = journal, doi = doi
  ))
  slug <- gsub("[^A-Za-z0-9]+", "-", doi)
  write_report(rep, file.path(corpus_dir, paste0(slug, ".csv")))
}

invisible(Map(make_sheet,
  n_reported = c(4, 9, 14, 6, 11),
  title      = paste("Study", 1:5),
  journal    = c("J Alpha", "J Alpha", "J Beta", "J Beta", "J Beta"),
  doi        = paste0("10.1234/study", 1:5)))

list.files(corpus_dir)
```

Now read them:

```{r}
corpus <- read_instar(corpus_dir)
corpus
```

A corpus is a named list of reports underneath, so `lapply()`,
`Filter()`, `[` and friends all work on it. `summary()` gives one row per
sheet:

```{r}
summary(corpus)[, c("study", "title", "journal", "reported", "percent_reported")]
```

### When a file will not read

Pointed at a real directory you will meet malformed files: a deleted
column, a semicolon-delimited export, a stray spreadsheet someone left
in the folder. `read_instar()` does not stop on those. It warns, skips
them, and records what went wrong.

```{r}
writeLines(c("this,is,not", "an,instar,sheet"),
           file.path(corpus_dir, "junk.csv"))

corpus <- suppressWarnings(read_instar(corpus_dir))
attr(corpus, "failed")
```

Getting the other five sheets plus a list of what to go and fix is more
useful than getting nothing.

### Framework versions

Every sheet records the version it was completed against. A corpus that
mixes versions is warned about, because coverage across them is not
comparable: an item missing from half the sheets because it did not exist
yet is not the same as an item those studies failed to report.

```{r}
table(summary(corpus)$version, useNA = "ifany")
```

If you do have a mix, split before auditing:

```{r, eval = FALSE}
s <- summary(corpus)
v1 <- corpus[s$version == "1.0"]
instar_audit(v1)
```

## Auditing

`instar_audit()` takes the corpus, or the path directly:

```{r}
audit <- instar_audit(corpus)
audit
```

`summary()` gives one row per framework item:

```{r}
head(summary(audit), 4)
```

The columns worth knowing: `applicable` is the denominator, excluding
studies that marked the item not applicable, and `percent_reported` is
`reported / applicable`. An item that genuinely does not apply to a study
is never counted against it.

Which items does this literature handle worst?

```{r}
s <- summary(audit)
head(s[order(s$percent_reported), c("item", "reported", "applicable",
                                    "percent_reported")], 5)
```

### Grouping

Any study-level column can be a grouping variable. For a corpus read from
sheets that means anything the reserved rows carry — `journal`, `doi`,
`version` — plus `folder`, the directory each sheet came from.

```{r}
by_journal <- summary(audit, by = "journal")
subset(by_journal, item_id == "subjects_taxon")
```

`folder` is the useful one when a corpus is filed into directories that
mean something. Read recursively and the structure comes through:

```{r, eval = FALSE}
corpus <- read_instar("corpus/", recursive = TRUE)
summary(instar_audit(corpus), by = "folder")
```

### Per-study coverage

`audit$studies` is one row per study, which is what you want for
questions about papers rather than items:

```{r}
stats::aggregate(percent_reported ~ journal, data = audit$studies,
                 FUN = stats::median)
```

### Plotting

```{r, fig.width = 7, fig.height = 6, fig.alt = "Bar chart of reporting coverage per framework item"}
plot(audit)
```

`autoplot()` returns the ggplot object instead of drawing it, so you can
modify it. `plot(audit, by = "journal")` facets.

## Studies with no sheet

Which is every study published before the framework existed. Score the
papers into a wide table: one row per paper, one column per `item_id`.

```{r}
scores <- data.frame(
  doi            = c("10.1/a", "10.1/b", "10.1/c", "10.1/d"),
  journal        = c("J Alpha", "J Alpha", "J Beta", "J Beta"),
  year           = c(2024, 2025, 2025, 2026),
  subjects_taxon = c("Y", "Y", "Y", "Y"),
  subjects_n     = c("Y", "N", "N", "Y"),
  env_housing    = c("Y", "Y", "NA", "N"),
  env_field      = c("NA", "NA", "Y", "-"),
  stringsAsFactors = FALSE
)

audit2 <- audit_from_matrix(scores, id = "doi")
summary(audit2)[, c("item", "reported", "applicable", "percent_reported")]
```

Three things are happening there worth spelling out.

**Cell values are read leniently**, because scoring sheets are made by
people. `Y`, `yes`, `TRUE`, `1` and `C` mean reported; `N`, `no`,
`FALSE` and `0` mean not; `NA`, `N/A`, `-` and blanks mean not
applicable. Anything unrecognised warns rather than being silently
counted.

**Non-item columns become study metadata.** `journal` and `year` are not
`item_id`s, so they ride along and are immediately available for
grouping:

```{r}
subset(summary(audit2, by = "year"), item_id == "subjects_n")
```

**Items absent from the table are left out**, not counted as unreported.
The message above says which. Recording an unscored item as a failure to
report would assert something the scoring never checked.

### Choosing the identifier

`id` should name a column that identifies every row. If it does not,
you get told rather than finding out later:

```{r}
incomplete <- data.frame(
  doi = c("10.1/a", "", ""),          # DOI extraction failed for two
  subjects_taxon = c("Y", "N", "Y"),
  stringsAsFactors = FALSE
)
audit3 <- suppressMessages(audit_from_matrix(incomplete, id = "doi"))
audit3$n
```

That warns about the blank ids and still counts three studies. It is a
real case: in the survey reported in the INSTAR paper, DOI extraction
failed for 25 of 150 PDFs, and keying on DOI would have collapsed them
into one repeated blank identifier. The fix was to key on a per-file
slug instead and keep DOI as metadata.

## A note on what this is for

The point of depositing the sheet as structured text rather than only as
a figure is that the second direction becomes possible at all. Once
sheets are routinely deposited, questions like which items a field
handles worst, whether reporting differs between laboratory and field
work, or whether it improves after a journal adopts the framework, are a
matter of reading files rather than of re-reading papers.

```{r, include = FALSE}
unlink(corpus_dir, recursive = TRUE)
```
