---
title: "Getting started with instarreport"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with instarreport}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>",
                      fig.width = 8.5, fig.height = 11, dpi = 100)
```

`instarreport` is the R implementation of **INSTAR** (INvertebrate
Standards for Treatment And Reporting), an 18-item reporting framework
for invertebrate welfare in research. It produces a standardised,
one-page summary figure for a paper in the spirit of the PRISMA and
ROSES flow diagrams. Each cell of the figure carries the substantive
content for the corresponding framework item, not just a tick-box.

```{r setup}
library(instarreport)
```

## The INSTAR framework

The package ships with the 18-item INSTAR framework as the exported
data frame `instar_items`.

```{r}
instar_items
```

Items are grouped into five welfare domains (Mellor *et al.* 2020, with
"affective state" in place of "mental state") and three cross-cutting
foundations. End-of-study disposition lives within the Health welfare
domain.

## Filling out the items

Three workflows, all producing the same `items` data frame that
`instar_report()` consumes.

### The INSTAR.csv sheet

The usual route, and the one the framework is designed around. Write the
blank sheet, fill in the `report` column, read it back.

```{r, eval = FALSE}
write_template("INSTAR.csv")
# ...fill the `report` column in Excel, Numbers, or any editor...
items <- read_items("INSTAR.csv")
```

One row per reporting item, carrying its `domain`, `item`, `description`
and applicability flags, with an empty `report` column at the far right
to type into. Reserved rows at the top hold a usage note, the framework
version, and the paper's own details, so a completed file explains
itself and identifies both the study it belongs to and the framework it
was completed against. That also means `instar_report()` can take it
whole:

```{r, eval = FALSE}
report <- instar_report(read_items("INSTAR.csv"))
```

A copy ships with the package:

```{r}
system.file("extdata", "INSTAR.csv", package = "instarreport")
```

The same rows are available as a formatted spreadsheet (`INSTAR.xlsx`)
for anyone who would rather not look at a CSV. Save it back out as CSV to
read it into R.

### Interactive prompt

The most forgiving workflow for first-time users. `instar_fill()` walks
through each item, showing its name and description, and prompts for a
value. It can save partway and resume later.

```{r, eval = FALSE}
items <- instar_fill(save_to = "INSTAR.csv")
```

At any prompt, type the value, or one of: `[enter]` to keep current,
`NA` for not applicable, `skip` to leave blank, `back` to step back,
`save` to write progress, `quit` to stop.

To resume:

```{r, eval = FALSE}
items <- instar_fill(read_items("INSTAR.csv"),
                    save_to = "INSTAR.csv")
```

To edit a single item afterwards (no need to know `item_id`s — pass no
`item_id` and you'll be shown a numbered menu):

```{r, eval = FALSE}
items <- instar_edit(items)
items <- instar_edit(items, "subjects_n")   # or by id
```

### Programmatic

Useful in scripts and pipelines.

`instar_set()` takes `item_id = "what the study reports"` pairs, as many
as you like at once:

```{r, eval = FALSE}
items <- instar_set(
  instar_template(),
  subjects_taxon = "Bombus terrestris (worker female); morphology + COI",
  subjects_n     = "n = 80; 72 analysed",
  env_field      = NA,       # does not apply
  proc_anaesthesia = NA
)
```

A string reports the item, `NA` marks it not applicable, and `""` leaves
it unreported. The same three states as the `report` column of the sheet.

Set `value` directly and it will not work — and will not tell you so.
`status` is the single source of truth, and `instar_report()` blanks
`value` for any item whose status is not `"reported"`, so a table filled
in that way renders as an entirely unreported study:

```{r, eval = FALSE}
# Does nothing useful: status is still "not_reported", so the text is
# dropped when the report is built.
items$value[items$item_id == "subjects_taxon"] <- "Bombus terrestris"
```

## Checking progress

```{r}
# A part-filled example to look at.
items <- read_items(
  system.file("extdata", "example_items.csv", package = "instarreport")
)
print(items)
```

prints all 18 items grouped by domain, marking each as reported (`*`),
not applicable (`-`), or blank (`o`), with the current value shown
underneath.

Every item carries a `status`, which is the single source of truth:

| `status`           | Meaning                         | Renders as             |
|--------------------|---------------------------------|------------------------|
| `"reported"`       | The study reports this item     | The text in `value`    |
| `"not_reported"`   | The study is silent on it       | *Not reported*, muted  |
| `"not_applicable"` | It does not apply to this study | *Not applicable*, grey |

`value` carries substantive content only, and is `NA` whenever `status`
is not `"reported"`, so the two can never disagree. Reading a CSV with no
`status` column derives it from `value`: blanks become `"not_reported"`,
and `"NA"` or `"N/A"` are honoured as shorthand for `"not_applicable"`.

## Build the report

`instar_report()` returns *data*, not a plot: the paper metadata, the
resolved item table, and a coverage summary. Rendering is a separate
step, so you can compute on a report before (or without) drawing it.

```{r}
report <- instar_report(
  items,
  paper = list(
    title   = "Field collection and laboratory choice trials in Bombus terrestris",
    authors = "Smith, Jones, & Lee (2025)",
    journal = "Behavioral Ecology, 36(4): 451-462"
  )
)

report
```

`summary()` gives one row per framework item, with its domain, group, and
status. This is the handle to use for anything programmatic.

```{r}
head(summary(report))

# Which items did this study not report?
subset(summary(report), status == "not_reported", select = c(domain, item))
```

## Render the figure

`plot()` draws the standardised figure.

```{r, fig.alt = "Example welfare reporting figure"}
plot(report)
```

If you want to modify the figure before rendering, `autoplot()` returns
the underlying patchwork composition instead of drawing it:

```{r, eval = FALSE}
autoplot(report) + patchwork::plot_annotation(caption = "Figure S1")
```

## Save it

A finished report has two deposit artifacts: the completed sheet and the
figure. The naming convention is that `write_` produces text and `save_`
produces figures.

```{r, eval = FALSE}
write_report(report, "INSTAR.csv")   # the completed sheet
save_figure(report,  "INSTAR.pdf")   # the figure
save_figure(report,  "INSTAR.png", dpi = 300)
```

`write_report()` writes the same file shape an author would have filled
in by hand, with the paper's details in the reserved rows, so it reads
straight back through `read_items()`. This is the machine-readable
artifact worth depositing; the figure is for human readers.

`save_figure()` renders the report and writes it via `ggplot2::ggsave()`,
choosing a compact page height from the figure's natural content size
(white background, 300 dpi for raster formats). Pass an explicit `height`
to override.

## Many studies at once

Everything above describes one study. Once sheets are being deposited
alongside papers, a set of them is a corpus, and the same machinery
summarises a literature rather than a study.

`read_instar()` is the loader. Point it at a file, a directory, or a
vector of either; it works out the format from the extension and always
returns an `instar_corpus`, whether it found one sheet or two hundred.

```{r, eval = FALSE}
corpus <- read_instar("supplements/")
corpus
#> <instar_corpus>
#>   45 sheets, INSTAR v1.0
#>   median coverage 71% (range 22-100%)
#>   2 files failed to read; see attr(., "failed")
```

A file it cannot read is reported and skipped rather than aborting the
run, with the error kept in `attr(corpus, "failed")`. In a corpus of any
size some fraction of files will be malformed, and the useful behaviour
is to get the other 199 and a list of what to go and fix.

Sheets are named by DOI where the reserved rows carry one, and by file
name otherwise. A corpus is a plain named list underneath, so `lapply()`,
`Filter()` and `[` all work on it as you would expect, and `summary()`
gives one row per sheet.

`instar_audit()` computes coverage per framework item across the corpus:

```{r, eval = FALSE}
audit <- instar_audit(corpus)      # or instar_audit("supplements/")

summary(audit)                     # one row per item
summary(audit, by = "journal")     # ...per item per journal
plot(audit)
```

Not-applicable items stay out of the denominator, per study and per
item, so an item that genuinely does not apply is never counted against
a study.

If a corpus is filed into folders that mean something, read it
recursively and the folder survives as a grouping variable:

```{r, eval = FALSE}
corpus <- read_instar("corpus/", recursive = TRUE)
summary(instar_audit(corpus), by = "folder")
```

### Studies that never filled in a sheet

Which is every study published before the framework existed. Score them
into a wide table instead, one row per paper and one column per
`item_id`:

```{r, message = FALSE}
scores <- data.frame(
  doi            = c("10.1234/a", "10.1234/b", "10.1234/c"),
  journal        = c("J Alpha", "J Beta", "J Beta"),
  subjects_taxon = c("Y", "Y", "Y"),
  subjects_n     = c("Y", "N", "N"),
  env_field      = c("NA", "Y", "-"),
  stringsAsFactors = FALSE
)

audit <- audit_from_matrix(scores, id = "doi")
summary(audit)
```

Columns matching an `item_id` are treated as items; every other column
is carried through as study metadata, so `journal` here is immediately
available to `summary(by = )` without building a lookup table. Cell
values are read leniently, because scoring sheets are made by people:
`Y`, `yes`, `TRUE`, `1` and `C` all mean reported, `N`, `no`, `FALSE`
and `0` mean not reported, and `NA`, `N/A`, `-` and blanks mean not
applicable.

Framework items absent from the matrix are left out of the audit rather
than counted as unreported, since the scoring never checked them.

### Framework versions

Each sheet records the INSTAR version it was completed against. Reading
a corpus that mixes versions warns, because coverage is not comparable
across them: an item missing from half the sheets because it did not
exist yet is not the same as an item those studies failed to report.

```{r, eval = FALSE}
table(summary(corpus)$version)
```

## Web tool

For a clickable interface, run:

```{r, eval = FALSE}
instar_app()
```

The Shiny app lets you fill in each framework item interactively with a
live preview, or upload a filled `INSTAR.csv` to render it. It downloads
the figure (PDF or PNG) and the completed `INSTAR.csv`. A hosted copy is
at <https://instar-statement.org/app/>, needing no installation.
