---
title: "Getting started with instarreport"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with instarreport}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>",
                      fig.width = 8.5, fig.height = 11, dpi = 100)
```

`instarreport` is the R implementation of **INSTAR** (INvertebrate
Standards for Treatment And Reporting), an 18-item reporting framework
for invertebrate welfare in research. It produces a standardised,
one-page summary figure for a paper in the spirit of the PRISMA and
ROSES flow diagrams. Each cell of the figure carries the substantive
content for the corresponding framework item, not just a tick-box.

```{r setup}
library(instarreport)
```

## The INSTAR framework

The package ships with the 18-item INSTAR framework as the exported
data frame `instar_items`.

```{r}
instar_items
```

Items are grouped into five welfare domains (Mellor *et al.* 2020, with
"affective state" in place of "mental state") and three cross-cutting
foundations. End-of-study disposition lives within the Health welfare
domain.

## Three ways to fill out the items

Pick whichever fits how you work. All three produce the same kind of
`items` data frame that `instar_report()` consumes.

### Interactive prompt

The most forgiving workflow for first-time users. `instar_fill()` walks
through each item, showing its name and description, and prompts for a
value. It can save partway and resume later.

```{r, eval = FALSE}
items <- instar_fill(save_to = "my_study.csv")
```

At any prompt, type the value, or one of: `[enter]` to keep current,
`NA` for not applicable, `skip` to leave blank, `back` to step back,
`save` to write progress, `quit` to stop.

To resume:

```{r, eval = FALSE}
items <- instar_fill(read_items("my_study.csv"),
                    save_to = "my_study.csv")
```

To edit a single item afterwards (no need to know `item_id`s — pass no
`item_id` and you'll be shown a numbered menu):

```{r, eval = FALSE}
items <- instar_edit(items)
items <- instar_edit(items, "subjects_n")   # or by id
```

### CSV template

Easier when you want to hand the form to a collaborator, or when items
are spread across multiple people. Write a blank template, fill it in in
Excel (or any spreadsheet or text editor), then load it back.

```{r, eval = FALSE}
write_template("my_study.csv", study_type = "field")
# ...edit the value column...
items <- read_items("my_study.csv")
```

The template CSV has one row per item with `item_id`, `item`, `domain`,
`description`, and a blank `value` column. The `item`, `domain`, and
`description` columns are in the file as on-page reminders of what each
item asks for. They are ignored on load, so only `item_id` and `value`
are strictly required.

A bundled blank template is also available as a static file you can
hand-copy:

```{r}
template_path <- system.file("extdata", "template_items.csv",
                             package = "instarreport")
template_path
```

And a filled example to look at:

```{r}
example_path <- system.file("extdata", "example_items.csv",
                            package = "instarreport")
items <- read_items(example_path)
head(items)
```

### Programmatic

Useful in scripts and pipelines.

```{r, eval = FALSE}
items <- instar_template()

# Reporting an item means setting both `value` and `status`.
items$value[items$item_id == "subjects_taxon"] <-
  "Bombus terrestris (worker female); morphology + COI"
items$status[items$item_id == "subjects_taxon"] <- "reported"

# Items that do not apply to the study:
items <- instar_na(items, c("env_field", "proc_anaesthesia"))
```

A small helper keeps this tidy if you are filling many items at once:

```{r, eval = FALSE}
report_item <- function(items, id, text) {
  items$value[items$item_id == id]  <- text
  items$status[items$item_id == id] <- "reported"
  items
}

items <- report_item(items, "subjects_n", "n = 80; 72 analysed")
```

## Checking progress

```{r, eval = FALSE}
print(items)
```

prints all 18 items grouped by domain, marking each as reported (`*`),
not applicable (`-`), or blank (`o`), with the current value shown
underneath.

Every item carries a `status`, which is the single source of truth:

| `status`           | Meaning                         | Renders as             |
|--------------------|---------------------------------|------------------------|
| `"reported"`       | The study reports this item     | The text in `value`    |
| `"not_reported"`   | The study is silent on it       | *Not reported*, muted  |
| `"not_applicable"` | It does not apply to this study | *Not applicable*, grey |

`value` carries substantive content only, and is `NA` whenever `status`
is not `"reported"`, so the two can never disagree. Reading a CSV with no
`status` column derives it from `value`: blanks become `"not_reported"`,
and `"NA"` or `"N/A"` are honoured as shorthand for `"not_applicable"`.

## Build the report

`instar_report()` returns *data*, not a plot: the paper metadata, the
resolved item table, and a coverage summary. Rendering is a separate
step, so you can compute on a report before (or without) drawing it.

```{r}
report <- instar_report(
  paper = list(
    title   = "Field collection and laboratory choice trials in Bombus terrestris",
    authors = "Smith, Jones, & Lee (2025)",
    journal = "Behavioral Ecology, 36(4): 451-462"
  ),
  items = items
)

report
```

`summary()` gives one row per framework item, with its domain, group, and
status. This is the handle to use for anything programmatic.

```{r}
head(summary(report))

# Which items did this study not report?
subset(summary(report), status == "not_reported", select = c(domain, item))
```

## Render the figure

`plot()` draws the standardised figure.

```{r, fig.alt = "Example welfare reporting figure"}
plot(report)
```

If you want to modify the figure before rendering, `autoplot()` returns
the underlying patchwork composition instead of drawing it:

```{r, eval = FALSE}
autoplot(report) + patchwork::plot_annotation(caption = "Figure S1")
```

## Save it

```{r, eval = FALSE}
instar_save(report, "fig_S1_welfare_reporting.pdf")
instar_save(report, "fig_S1_welfare_reporting.png", dpi = 300)
```

`instar_save()` renders the report and writes it via
`ggplot2::ggsave()`, choosing a compact page height from the figure's
natural content size (white background, 300 dpi for raster formats).
Pass an explicit `height` to override.

## Web tool

For a clickable interface, run:

```{r, eval = FALSE}
instar_app()
```

The Shiny app lets you fill in each framework item interactively, preview
the figure, and download the PDF or PNG.
